Tail associations and their impact on extinction risk

### 2019-11-01 MainText.Rmd
done with discussion for now

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
instead of attempting mathematical analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination other researchers. Some material in the Discussion outlines some possible next steps.
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

<!--***DAN: stopped here-->

@@

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the oposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across large areas. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. However, our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates.

<!--***DAN: May be scope for a discussion para linking what is known about
how ECEs are changing to what effect that may have on how tail assiciations
in environmental variables might be changing, and what influence that may have
on extinction risk. If little is known, then this might be a good topic for 
future work.-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and specific events can now be attributed with confidence to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechansms that were explored in this paper.

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
<!--***DAN: Dan to elaborate and make more specific/precise-->
We found a left tail association in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable 
unless the tail association was at its 
extreme (Fig. \ref{fig_extrisk_agestr}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the 
stochastic log(growth rate) in an iid environment.
The approximation considered growth rate being 
dependent on the most dominant eigenvalue of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused 
by environmental stochasticity - which was based
on the linear covariance measure between any 
two matrix elements, not on their tail association properties. Therefore,
following @tuljapurkar1990lecture, the population growth 
rate in an iid environment could be approximated
as same for our model (Eq. \ref{eq_age_str}) with 
left-tail or right-tail association.

<!--***DAN: Dan to maybe add another para on the fact that I thnk Tulja's approx should also
work for our spatial, non-dd model (check first, Dan), but we see a strong effect here,
so not clear why. And also suggests the possibilty that for some age-structured models,
tail associations will make a bigger difference then they did in our case. Maybe somehow
get to the suggestion that someone needs to revisit Tulja's approx and the analytic
approximation of extinction risk of Lande and Orzack looking at tail associations? Maybe
this and the previous para need to be combined, or broken up in a different way.-->

<!--***DAN: maybe you want to have a Discussion para on copulas. It could start with
some text about ranks being the proper way to study associations, and then could go
on to link to copulas and cite BIVAN-->

<!--***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results. Interestingly, 
Tulja's approximation may also apply to the spatial case, at least when 
all locations can be reached from all other locations via dispersal. But the 
approximation seems poor in that case. Need to think about it more, and maybe
need more research.-->

<!--Intution/ possible explanation for expected results/ hypo?-->

<!--future study: worthy for PVA, IUCN red list:
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula? Multisite age str. model with dispersal, directional change
   in spatio-temporal variation for vital rates.
3. Tdep. with -ve correlation between vital rates?
4. extinction risk how propagated through a food web?
5. Not sure : see McCarthy2000: Most previous models of metapopulation viability have either ignored
correlated extinction of patches, or have included uniform correlations, with the correlation between 
nearby patches being the same as the correlation between distant
patches (Hanski 1994; Tilman et al. 1994; Lindenmayer et al. 1995; McCarthy
1996). 
-->
Although we demonstrated that tail association between vital rates of an age-structured model can
influence extinction risk, substantial work remains to determine the importance of this effect in the 
following scenarios. First, we demonstrated the effect of the tail association on extinction risk 
with a basic age-structured (two age stages) model where we considered the total semelparous 
population would go extinct if there would be only one female adult and one egg. But in reality population 
viability analysis (PVA) deals with a little bit higher threshold with several age, stage or size class and a variety 
of populations including semelparous and iteroparous [@morris2002quantitative]. Future studies should include 
tail association effect on extinction risk 
for a multiple age-, size- and stage-structured model and estimate extinction risk 
for each age, size and stage class separately 
so that we can identify (via multisite-PVA) which life-stage or demographic process 
to target for population management. An earlier study by @doak1994modeling showed PVA models with correlated 
demographic rates indicated a marked difference in predicting confidence interval for 
future population size compared with the model which did not include correlation. But when 
such correlation cannot be estimated directly, more generally where the
forms of statistical dependence among the model's parameters are unknown, making incorrect
assumptions about correlations and dependencies may lead to severe over-estimation 
or under-estimation of extinction risks [@ferson1995correlations]. We believe modeling 
dependencies with copula could help in this case. 
<!--***DAN: the ideas in this para up to here (taking out the wrong statements) might 
follow on well from a para based on the notes I wrote above on Tulja's approx.-->
Second, we did not consider multiple sites of 
age-structured metapopulation which could be linked via dispersal. So, the possible future investigation 
would be to explore the viability for the metapopulations if vital rates (e.g. fecundity or 
survival of different stage class) across sites showed specific tail association 
with spatially correlated extreme climatic events. 
<!--***DAN: a reasonable point, above-->
Third, we considered the tail association 
between positively correlated vital rates, but what would be the extinction risk 
and viability of a population if demographic rates were negatively correlated with the asymmetric tail association. 
<!--***DAN: a good point, above-->
We could consider this scenario to model extreme events that kill adults and allow more successful
recruitment of seeds or other propagules. Fourth, following the aforementioned idea, it would be worthy 
to model more complex prey-predator model where asymmetric tail association could feature the
interspecific-dependence. 
<!--***DAN: mentioning interactions strikes me as a step too far-->
An earlier study [@carnicer2011widespread] showed extreme events
(i.e., heatwaves and droughts) increased mortality and defoliation in the southern European forest and its 
dependent herbivorous insects and fungi through trophic linkage. <!--ECEs can even truncate 
multiyear insect outbreak dynamics [@esper20061200].--> We believe copula (maybe a multidimensional one) has 
a lot more potential to estimate the risk of cascading extinction in model food webs. <!--should I 
mention about 
match-mismatch hypothesis leads to extinction? our phenology paper in prep.--> Fifth, we considered 
our metapopulation patches were equispaced on a linear chain with equal global dispersal 
rate between patches. But in realistic network models, as inter-patch distances 
can be different, the dispersal rate should depend on distance: dispersal is possible at higher 
rates between nearby patches than the patches which are far away. It is worthy to study 
metapopulation viability with non-uniform spatial correlation (i.e. asymmetric tail association) in a 
network with different clusters having various dispersal rates.
<!--***DAN: might also mention how this relates with geog of sync of environmental vars, linking to
Jon's paper-->

>>>>

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
for both Q1 and Q2 instead of attempting mathematical 
analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination other researchers. Some material in the Discussion 
outlines some possible next steps.
<!--***DAN: make sure you do this-->
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

@@

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the oposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across large areas. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates. Our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model (but see below).

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechansms we explored in this paper. 
If $X_i(t)$ denotes the value of an environmental variable (e.g., a summer maximum 
temperature, or winter minimum temperature, spring rainfall, etc.) 
in location $i$ in year $t$,
@ummenhofer2017extreme provide a diagram of how ECEs may be increasing in frequency 
and intensity due to changes in the distribution of $X_i(t)$ such as changes
in the mean, variance, or skewness of this distribution. Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. Documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE given that one of them is so
influenced should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even through Pearson correlation
appears to inducate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends on both marginal information and association information.
It is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of  
variables, rather than reflecting true changes in association.
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018, @Valpine2014, @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, and how such 
associations may be changing.

<!--Tuljapurkar approximation-->
Our results suggest that tail associations and asymmetries of tail associations
may have limited influence on the extinction risk of 
stage-structured, non-spatial models. Although we believe more research is 
certainly still needed
as to whether this conclusion holds more generally than the model we considered
(see below), if it does hold 
generally, it may be possible to obtain a better understanding of 
the result using a well-known approximation
of Tuljapurkar [@tulja1982; @tulja1990] for the long-term stochastic growth rate.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extintion risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations.
We also note that this reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

<!--***DAN: decided the below was a step too far
Finally, it seems likely that the 
stochastic matrix modelling framework, and Tuljapurkar's approximation,
should apply to the density-independent metapopulation model of 
@Ghosh_copula, which was comprised of Lewontin-Cohen models linked 
by dispersal. Nevertheless asymmetry of tail associations made a 
big difference to extinction risk for that model [@Ghosh_copula].
We have not examined whether Tuljapurkar's approximation is
accurate in that spatially explicit case.

***DAN: Dan to maybe add another para on the fact that I thnk Tulja's approx should also
work for our spatial, non-dd model (check first, Dan), but we see a strong effect here,
so not clear why. And also suggests the possibilty that for some age-structured models,
tail associations will make a bigger difference then they did in our case. Maybe somehow
get to the suggestion that someone needs to revisit Tulja's approx and the analytic
approximation of extinction risk of Lande and Orzack looking at tail associations?

***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results. Interestingly, 
Tulja's approximation may also apply to the spatial case, at least when 
all locations can be reached from all other locations via dispersal. But the 
approximation seems poor in that case. Need to think about it more, and maybe
need more research.-->

<!--future study: worthy for PVA, IUCN red list:
1. Identify which life stage or demographic process as management target?
2. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula? Multisite age str. model with dispersal, directional change
   in spatio-temporal variation for vital rates.
3. Tdep. with -ve correlation between vital rates?
4. extinction risk how propagated through a food web?
5. Not sure : see McCarthy2000: Most previous models of metapopulation viability have either ignored
correlated extinction of patches, or have included uniform correlations, with the correlation between 
nearby patches being the same as the correlation between distant
patches (Hanski 1994; Tilman et al. 1994; Lindenmayer et al. 1995; McCarthy
1996). 
-->

<!--future ideas-->
In our opinion, additional research would be worthwhile on the 
impact of tail associations on growth rates and 
extinction risk under additional modelling scenarios not considered
in this paper.   
First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, relating to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be well be associated with,
for instance, $r_2$, whereas lower values of $r_1$ may not be strongly
associated with $r_2$ because reductions/increases in $r_1$ can
then be traded off against increases/reductions in *either* 
$r_2$ or $r_3$. Second, for our metapopulation models,
we assumed habitat patches were equispaced on a transect or
were all equally mutually accessible through dispersal.
Furthermore, associations between environmental variables in different
patches were assumed to be spatially homogeneous. In real metapopulations,
neither of these assumptions holds. Connecting the ideas of tail association that
we have developed in this study with the ideas on the geography of synchrony
of @walteretal2017 seems like a promising future direction.
